A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-10-16 doc/large_strain.tex

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound.
    The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of the iterative solver, so-called matrix-free methods are
    widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.

    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyper-elastic tangent operator numerically.
    In order to improve the convergence behavior of iterative solvers, we also propose a way to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    Our implementation employs MPI {\color{red}and Intel Threading Building Blocks parallelization} and SIMD vectorization.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
  \end{abstract}

@@

The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of the iterative solver, so-called matrix-free methods are
widely adopted in fluid mechanics community, where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}. Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
A possible reason is that the tangent operators (coming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used in fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case. Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution. FEM with linear or quadratic elements, that avoid locking, are often adopted in the solid mechanics community. Higher order elements are rarely employed. However, we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

In this work, we investigate different implementations of the finite-strain hyper-elastic tangent operator with respcet to their computational cost.
In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

The paper is organized as follows. In Section \ref{sec:theory}, we briefly introduce the partial differential equations used in  finite-strain solid mechanics. Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches. In Section \ref{sec:gmg}, we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity. The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}. The results are summarized in Section \ref{sec:summary}.

@@

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
We assume that the deformation map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian has to be positive $J = \det \gz F > 0$. The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}

%
% strain measures
%
In what follows, we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$. Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$. We will also use the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T = \gz V^2$.

For conservative systems without volumetric load, the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
\end{align}
Following \cite{Wriggers2008}, this can be simplified to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the the linearization of the deformation gradient $\gz F$.

@@

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to obtained as solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products.
%
The matrix-vector product $A x$ can be expressed by
\begin{align*}
 (Ax)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &= \sum_K\sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q
\end{align*}
where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds
\begin{align*}
 \tilde{a}_K(N_i,N_j)(\gz x_q) = N_i(\gz x_q)N_j(\gz x_q).
\end{align*}
Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
\begin{lstlisting}
template<int k, int q>
void evaluate_dof_vector_in_all_quadrature_points
  (const double u[k][k][k],
   const double phi[k][q],
   double return_value[q][q][q])
{
  double tmp1[k][k][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int i2=0; i2<k; ++i2)
      for (int q3=0; q3<q; ++q3)
        for (int i3=0; i3<k; ++i3)
          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];

@@

  for (int q2=0; q2<q; ++q2)
    for (int q3=0; q3<q; ++q3)
      for (int q1=0; q1<q; ++q1)
        for (int i1=0; i1<k; ++i1)
  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
}
\end{lstlisting}
in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II}\cite{dealII90} library.
In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II} the integration is done over the deformed configuration. Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
The efficiency of standard iterative solvers deteriorate for large systems of linear equations due to incresing iteration counts under mesh refinement.
On the other hand, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts independent of the number of mesh refinements by smoothing the defect on a hierarchy of levels. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.

>>>>

  \begin{abstract}
    The performance of finite element solvers on modern computer architectures is typically memory bound.
    The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
    In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
    widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.

    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic tangent operator numerically.
    In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
  \end{abstract}

@@

The performance of finite element solvers on modern computer architectures is typically memory bound.
The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
In order to improve the performance of iterative solvers, so-called matrix-free methods,
where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
are widely adopted in the fluid mechanics community.
Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a masters thesis produced by a student of two of the authors \cite{Mentler2017}.
One possible reason is that the tangent operators (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution.
Finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
Even higher order elements are rarely employed.
However, we note, even with linear FEM, that large scale computations with $10^{12}$ unknowns inevitably requires a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respcet to their computational cost.
In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
and employ them to define a geometric multigrid preconditioner.
Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.

The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics. Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches. In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity. The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}. The results are summarized in Section \ref{sec:summary}.

@@

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
We assume that the deformation map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second order unit tensor.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}

%
% strain measures
%
In what follows, we parameterize the material behavior using the right Cacuhy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
We will also use the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$ and the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T$.
Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
%
For conservative systems without the body force, the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
\end{align}
Following \cite{Wriggers2008}, this can be simplified to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of the inverse of the deformation gradient $\gz F^{-1}$.

@@

We now introduce a FE triangulation $\mathcal{B}^h_0$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the set of unknown degrees of freedom (DoF).
%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.
Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $A x$ can be expressed by
\begin{align*}
 (Ax)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &= \sum_K\sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q
\end{align*}
where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $q$.
For the sake of demonstration, let us assume that $A$ represents a mass matrix. In this case, it holds that
\begin{align*}
 \tilde{a}_K(N_i,N_j)(\gz x_q) = N_i(\gz x_q)N_j(\gz x_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently.
Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
\begin{lstlisting}
template<int k, int q>
void evaluate_dof_vector_in_all_quadrature_points
  (const double u[k][k][k],
   const double phi[k][q],
   double return_value[q][q][q])
{
  double tmp1[k][k][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int i2=0; i2<k; ++i2)
      for (int q3=0; q3<q; ++q3)
        for (int i3=0; i3<k; ++i3)
          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];

@@

  for (int q2=0; q2<q; ++q2)
    for (int q3=0; q3<q; ++q3)
      for (int q1=0; q1<q; ++q1)
        for (int i1=0; i1<k; ++i1)
  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
}
\end{lstlisting}
in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the consistent quadrature rule.
More information on techniques developed to increase the computational density can be found in \cite{kronbichler12,vos10}.
We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration. Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}}$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorate for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
